---
title: |
  | Introduction to Social Data Analytics
  | Class 22
author: ""
date: ""
output:
  beamer_presentation:
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Today: Plotting in ```R```

By the end of today's lecture, you should be able to:

- Create the following plots in R: barplot, histogram, boxplot, line plots, and scatter plots
- Recall how to generate tables and which plots require tables as inputs
- Add elements to plots: titles, axis labels, ablines, text, colors, etc.
- Interpret elements of plots after creating them (e.g. quartiles in box plots)

Open class22.R if you haven't already and fill-in as we go.

## On your own, load afghan.csv and explore the data
```{r, echo = FALSE} 
afghan <- read.csv("afghan.csv")
```

```{r, echo = TRUE, eval = TRUE}
names(afghan)
class(afghan$violent.exp.ISAF)
summary(afghan$violent.exp.ISAF)
```

## What's the difference between ```table``` vs. ```prop.table```

```{r, echo = TRUE, eval = TRUE}
ISAF.table <- table(afghan$violent.exp.ISAF, 
                    exclude = NULL)
ISAF.table
ISAF.ptable <- prop.table(table(afghan$violent.exp.ISAF,
                    exclude = NULL))
ISAF.ptable
```

## Create a barplot of the \textit{percent} victimized by ISAF

```{r, echo = TRUE, eval = FALSE}
barplot(ISAF.ptable,
        names.arg = c("No harm", "Harm", "Nonresponse"), 
        main = "Civilian victimization by the ISAF",
        xlab = "Response category",
        ylab = "Proportion of the respondents", 
        ylim = c(0, 0.7))
```

## Your barplot should look like this.

```{r, echo = FALSE, eval = TRUE}
barplot(ISAF.ptable,
        names.arg = c("No harm", "Harm", "Nonresponse"), 
        main = "Civilian victimization by the ISAF",
        xlab = "Response category",
        ylab = "Proportion of the respondents", 
        ylim = c(0, 0.7))
```

- Your turn! Create a barplot for ```afghan$violent.exp.taliban```

## Create a histogram of respondent ages
```{r, echo = TRUE, eval = FALSE}
hist(afghan$age, freq = FALSE, 
     ylim = c(0, 0.04), 
     xlab = "Age", 
     ylab = "Percent",
     main = "Distribution of Respondent Age")
```

## Your histogram should look like this
```{r, echo = FALSE, eval = TRUE}
hist(afghan$age, freq = FALSE, 
     ylim = c(0, 0.04), 
     xlab = "Age", 
     ylab = "Percent",
     main = "Distribution of Respondent Age")
```

- See if you can do the same for ```afghan$educ.years``` (Notice ```breaks()```)

## Suppose you want to add a vertical line though the median...

```{r, echo = TRUE, eval = FALSE}
# Add a vertical line at the median education level 
# using abline()
abline(v = median(afghan$educ.years))

# Add a text label "median" at (x, y) = (3, 0.5)
text(x = 3, y = 0.5, "median")

# Add a vertical line at the mean using lines()
lines(x = rep(mean(afghan$educ.years), 2), 
      y = c(-100, 1500))

```

- Try to add a text label "mean" in an appropriate place.

## Can we create a histogram for ```afghan$income```? Why or why not?

```{r, echo = TRUE}
summary(afghan$income)
class(afghan$income)
```

## Make a box plot of years of education separated by province
```{r, echo = TRUE, eval = FALSE}
boxplot(educ.years ~ province, 
        data = afghan, 
        main = "Education by Province", 
        xlab = "Province",
        ylab = "Years of Education")
```

## Which provinces are the most educated?
```{r, echo = FALSE, eval = TRUE}
boxplot(educ.years ~ province, 
        data = afghan, 
        main = "Education by Province", 
        xlab = "Province",
        ylab = "Years of Education")
```

- Make a boxplot of age separated by each district.

## Now load congress.csv and explore the data on your own 
```{r, echo = TRUE}
congress <- read.csv("congress.csv")
head(congress, 3)
```

## Subsetting has been done for you.
```{r, echo = FALSE, echo = FALSE}
rep <- subset(congress, subset = (party == "Republican"))
dem <- congress[congress$party == "Democrat", ] # alternative method

# Further subset the data by party and by 80th and 112th congress
rep80 <- subset(rep, subset = (congress == 80))
dem80 <- subset(dem, subset = (congress == 80))
rep112 <- subset(rep, subset = (congress == 112))
dem112 <- subset(dem, subset = (congress == 112))

# Assign axis labels and limits to avoid repetition
xlab <- "Economic liberalism/conservatism"
ylab <- "Racial liberalism/conservatism"
lim <- c(-1.5, 1.5)
```

- What is ```rep80```?
- What is ```dem112```?
```{r, echo = TRUE}
summary(rep80)
```

## Create a scatter plot demonstrating ideological division
```{r, echo = TRUE, eval = FALSE}
plot(1, type = "n", # Type "n" specifies no plotting
     xlim = lim, 
     ylim = lim, 
     xlab = xlab, 
     ylab = ylab,
     main = "80th Congress")
points(dem80$dwnom1, dem80$dwnom2, 
       pch = 16, col = "blue") # democrats
points(rep80$dwnom1, rep80$dwnom2, 
       pch = 17, col = "red") # republicans
text(-0.75, 1, "Democrats")
text(1, -1, "Republicans")
```

## Your scatter plot should look like this
```{r, echo = FALSE, eval = TRUE}
plot(1, type = "n", # Type "n" specifies no plotting
     xlim = lim, 
     ylim = lim, 
     xlab = xlab, 
     ylab = ylab,
     main = "80th Congress")
points(dem80$dwnom1, dem80$dwnom2, pch = 16, col = "blue") # democrats
points(rep80$dwnom1, rep80$dwnom2, pch = 17, col = "red") # republicans
text(-0.75, 1, "Democrats")
text(1, -1, "Republicans")
```

- Create the same plot for the 112th congress

## Now we create line plots showing ideology change over time
- First, let's generate vectors of median ideology vs time for each party
```{r, echo = TRUE}
# Calculate party median for each congress
dem.median <- tapply(dem$dwnom1, dem$congress, median)
rep.median <- tapply(rep$dwnom1, rep$congress, median)
```

## Now we can plot lines

```{r, echo = TRUE, eval = FALSE}
plot(names(dem.median), dem.median, 
     col = "blue", 
     type = "l",
     xlim = c(80, 115), 
     ylim = c(-1, 1), 
     xlab = "Congress",
     ylab = "Median ideological leaning of party")
lines(names(rep.median), rep.median, col = "red")
text(110, -0.6, "Democratic\n Party")
text(110, 0.85, "Republican\n Party")
```

## Does your plot look like this?

```{r, echo = FALSE, eval = TRUE}
plot(names(dem.median), dem.median, 
     col = "blue", 
     type = "l",
     xlim = c(80, 115), 
     ylim = c(-1, 1), 
     xlab = "Congress",
     ylab = "Median ideological leaning of party")
lines(names(rep.median), rep.median, col = "red")
text(110, -0.6, "Democratic\n Party")
text(110, 0.85, "Republican\n Party")
```
