---
title: |
  | Introduction to Social Data Analytics
  | Class 20
author: ""
date: ""
output:
  beamer_presentation:
    keep_tex: true
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Today: more practice with loops

By the end of today's lecture, you should be able to:

- Build loops that utilize the 'counter' (e.g. i) in three ways:
    - As a number for calculations
    - As a subset index
    - As an element of a vector (of strings or numbers)
- Construct ```while``` loops that may include conditional statements

Open class20.R if you haven't already. 


## Tools in your loop toolbox

- ```for```, ```if```, ```else```, and (soon) ```while```
- Using ```i``` as a number itself
- Using ```i``` to subset data
- Using ```i``` as an element of a vector

## The problem: no ```year``` vector

From pre class exercise: 

- Suppose I have data on the US population from 1790 - 1970
- But I don't have a ```year``` vector
```{r, echo = TRUE, eval = TRUE}
data(uspop)
summary(uspop)
n <- length(uspop)
```
- How could I generate a ```year``` vector using loops?

## Option 1: use ```i``` as a number itself
```{r, echo = TRUE, eval = FALSE}
for(i in 1:n){
  print(1780 + 10*i)
}
```

## Option 1: use ```i``` as a number itself
```{r, echo = TRUE, eval = TRUE}
for(i in 1:n){
  print(1780 + 10*i)
}
```


## Option 2: use ```i``` to subset data
```{r, echo = TRUE, eval = FALSE}
year <- rep(NA, n)
for(i in 1:n){
  year[i] <- 1780 + 10*i
  print(year[i])
}
```

## Option 2: use ```i``` to subset data
```{r, echo = TRUE, eval = TRUE}
year <- rep(NA, n)
for(i in 1:n){
  year[i] <- 1780 + 10*i
  print(year[i])
}
```

## Option 3: ```i``` as an element of a vector
```{r, echo = TRUE, eval = FALSE}
year <- seq(1790, 1970, 10)
for(i in year){
  print(i)
}
```
- This use is useful when ```i``` refers to an element of a vector of text strings

## Option 3: ```i``` as an element of a vector
```{r, echo = TRUE, eval = TRUE}
year <- seq(1790, 1970, 10)
for(i in year){
  print(i)
}
```

## Third use with a vector of text strings
```{r, echo = TRUE, eval = TRUE}
instructors <- c("Arushi", "Cameron", "Duy", 
                 "Mitch", "Zack")
for(i in instructors){
  print(paste(i,"is an instructor."))
}
```

## An example with ```for```, ```if```, and ```else```
- For each number ```i``` in 1:10, output whether ```i``` is even.
- To do this, we will use ```if``` statements within a ```for``` loop.
- First, let's introduce ```%%``` and ```paste()```:
```{r, echo = TRUE}
7 %% 5
remainder <- 7 %% 5
paste(remainder,"is the remainder")
```

## An example with ```for```, ```if```, and ```else```
Step 1:

- Write the ```if``` statements needed to test whether a number ```i``` is even
- If it is even, output the number and "is even."
- If it is odd, output the number and "is odd"

## An example with ```for```, ```if```, and ```else```
```{r, echo = TRUE, eval = FALSE}

  if(i %% 2 == 0){ 
    print(paste(i,"is even."))
  } else {
    print(paste(i,"is odd."))
  } 

```
- Now do Step 2: add the ```for``` loop so that the commands above are performed over all values in 1:10

## An example with ```for```, ```if```, and ```else```
```{r, echo = TRUE, eval = FALSE}
for(i in 1:10){
  if(i %% 2 == 0){ 
    print(paste(i,"is even."))
  } else {
    print(paste(i,"is odd."))
  } 
}
```

## An example with ```for```, ```if```, and ```else```
```{r, echo = TRUE}
for(i in 1:10){
  if(i %% 2 == 0){ 
    print(paste(i,"is even."))
  } else {
    print(paste(i,"is odd."))
  } 
}

```

## ***```while```*** Loops in ```R```

- The general form is: 
```{r, echo = TRUE, eval = FALSE}
while(logical test){
  do some stuff until logical test becomes false
} 
```
- For example:
```{r, echo = TRUE, eval = FALSE}
while(money > $0){
  eat(food)
  money <- money - cost(food)
} 
```
- When does the loop end? What happens if food is free?

## Example in ```R```
```{r, echo = TRUE}
x <- 0
limit <- 10
while(x <= limit){
  x <- x + 2
  print(x)
}
```
How many iterations does this loop go through?

## An example with ```while```, ```if```, and ```else```
```{r, echo = TRUE}
i <- 1
while(i <= 10){
  if(i %% 2 == 0){ 
    print(paste(i,"is even."))
  } else {
    print(paste(i,"is odd."))
  } 
  i <- i + 1
}

```

## Your turn!
- Work on class20.R and/or class19.R. 

Here are the new commands/operators we covered today:

- ```while```
- ```%%```
- ```paste```
