---
title: |
  | Introduction to Social Data Analytics
  | Class 19
author: ""
date: ""
output:
  beamer_presentation:
    keep_tex: true
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Today: loops

By the end of today's lecture, you should be able to:

- Construct conditional statements and for loops in ```R```
- Describe how loops can reduce coding necessary to accomplish data analysis
- Define "iteration" and give examples of how the "counter" can be used within a for loop
- Recall from the Excel lectures how to use the \texttt{if} operator and describe the syntax in \texttt{R}


Open class19.R if you haven't already. 

## ***```if```*** statements in ```R```
- The general form is: 
```{r, echo = TRUE, eval = FALSE}
if(logical test) {
  do some stuff when logical test is TRUE
} 
```
- For example:
```{r, echo = TRUE, eval = FALSE}
if(locked(door) == 1) {
  unlock(door)
} 
```
- As in Stata, the action only takes place when the if statement is true.

## Example in ```R```
```{r, echo = TRUE}
x <- sample(-1:1, 1) # random draw from {-1, 0, 1}
x
if (x > 0) {
  print("x is positive")
}
```

## Learn this...or ***```else```***!
- Use ```else``` to add conditions to an ```if``` statement: 
```{r, echo = TRUE, eval = FALSE}
if(logical test) {
  do some stuff when logical test is TRUE
} else {
  do some stuff when logical test is FALSE
}
```
- For example:
```{r, echo = TRUE, eval = FALSE}
if(locked(door) == 1) {
  unlock(door)
  open(door)
} else {
  open(door)
}
```

## Example in ```R```
```{r, echo = TRUE}
x <- sample(-1:1, 1)
x
if (x > 0) {
  print("x is positive")
} else { 
  print("x is negative")
}
```

## What about zero?
```{r, echo = TRUE}
x <- sample(-1:1, 1)
x
if (x > 0) {
  print("x is positive")
} else if (x < 0) { 
  print("x is negative")  
} else {
  print("x is zero")
}
```

## Takeaways from combining ```if``` and ```else```
- You do not specify the criterea for the ```else``` part
- These statements are \textit{exhaustive}: they check \alert{all} possible cases 
- Make sure the ```else``` shows up on the same line as the prior closed bracket

## What is a loop?

- A sequence of commands to be repeated
- You do various sorts of loops in your daily life
```{r, echo = TRUE, eval = FALSE}
for(i in 1:length(homework)){
  do(homework[i])
  homework[i] <- 0
}
if(sum(homework)) == 0 {
  watch(Netflix)
}
```
- In ```R```, they are helpful for tasks you would like repeated, where typing repeatedly would be cumbersome.

## ***```for```*** Loops in ```R```

- The general form is: 
```{r, echo = TRUE, eval = FALSE}
for(some set of things){
  do some stuff
} 
```
- For example:
```{r, echo = TRUE, eval = FALSE}
for(i in c(keys, phone, wallet)){
  check.on.person(i)
} 
```
- The loop \alert{iterates} through multiple rounds.
- How many iterations in the example?

## Example in ```R```
```{r, echo = TRUE}
x <- 0
for(i in 1:5){
  x <- x + 2*i
  print(x)
}
```
What is the value of ```x``` after the fifth iteration of the loop?


## Your turn: practice conditional statements and loops
Words of wisdom

- Loops are in general slow to compute
- Faster to use vector/matrix operations
- Try each problem with loop solutions to start

Here are the commands/operators we covered today:

- ```for, if, else```
- ```in, print, sample```
