---
title: |
  | Introduction to Social Data Analytics
  | Class 17
author: ""
date: ""
output:
  beamer_presentation:
    keep_tex: true
  'beamer_presentation: keep_tex: true': default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Today: logic and subsetting in ```R```

By the end of today's lecture, you should be able to:

- Write logic statements in R and identify the relevant Boolean operators
- Generate subsets of data using logic operators and \texttt{\$}

Open class17.R if you haven't already. 

## Any questions from the pre class exercise?

```{r, echo = TRUE, eval = TRUE}
"Triton" != "triton"
5<3 | FALSE
vec1 <- c(TRUE, TRUE, FALSE, TRUE); sum(vec1) > 3
```

## Creating lists and sequences
- Suppose you want a list of numbers increasing by 1
```{r, echo = TRUE}
list1 <- 1:5
list1
list1 > 3
(list1 > 1) & (list1 < 5)
```

## Now suppose you want a list of numbers increasing by 0.5
- Use the command ```seq(start, end, increment)```
```{r, echo = TRUE}
sequence1 <- seq(1,5,0.5)
sequence1
(sequence1 > 1) & (sequence1 < 5)
```
- With your partner, generate a sequence with six elements starting at 10 and ending at 20.

## Some "class" examples

```{r, echo = TRUE}
x = 5
class(x)
y = "My name is King Triton."
class(y)
z = TRUE
class(z)
```

## The \textit{class} function outputs the class of the object you insert

- Separate classes cannot in general be combined unless you convert them first
- We can \alert{coerce} objects to be recognized as different classes

## Objects: coercion with "as" 
```{r, echo = TRUE}
z.str <- c("1", "2", "602")
z.num <- as.numeric(z.str)
z.str
z.num
# z.str + c(1,1,1) will give an error
z.num + c(1,1,1)
```

## More "as" functions
```{r, echo = TRUE}
a_num <- c(1,2,3, 3498)
a_str <- as.character(a_num)
a_num
a_str
a_num + c(1,1,1,1)
# a_str + c(1,1,1,1) will give an error
```

## But, "is" it what you think?
```{r, echo = TRUE}
is.character(z.num)
is.character(z.str)
is.numeric(a_num)
is.numeric(a_str)
```

## How can we tell if something ```isTRUE```?
```{r, echo = TRUE}
isTRUE(TRUE)
isTRUE(FALSE)
isTRUE(T)
isTRUE(1)
```

## But be careful...

```{r, echo = TRUE}
as.integer(TRUE)
T <- 1
isTRUE(T)
```
- T is a predefined "global" that is equivalent to TRUE, but can be changed.

## Also...
```{r, echo = TRUE}
logic <- c(TRUE, TRUE, FALSE, TRUE)
mean(logic)
sum(logic)
```

## Coercion 
- Coercion is when you call a variable of the wrong type and R tries to fix it for you.
```{r, echo = TRUE}
vec2 <- c(1,2,3)
vec2
typeof(vec2)
class(vec2)
```

## But ...

```{r, echo = TRUE}
vec2[1] = "text"
class(vec2)
vec2
# or
TRUE == "TRUE"
```
- Both are examples of coercion.

## Subsetting
- ```[]``` is for subsets
- ```$``` is for extracting by name (so what you want must be named to be used...)
- What does ```resume$race[1:5]``` give us?
- What does ```resume[,"race"]``` give us?
- What does ```mean(resume$call[resume$race == "black"])``` give us?

## Using subset()
- ```R``` has a function that can help generate subsets of dataframes
```{r, echo = TRUE, eval = FALSE}
black <- subset(resume, select = c("firstname", "sex", 
        "call"), subset = (race == "black"))
```

## Here are the commands/operators we covered today:

- ```&, |, !, ==```
- ```:, seq()```
- ```class(), typeof()```
- ```as.numeric(), as.character()```
- ```is.numeric(), is.character(), isTRUE()```
- ```subset()```